
```{r stream-setup-octo}
#| code-summary: "Setup: packages, color palette"
suppressWarnings(suppressPackageStartupMessages({
  library(tidyverse)
  library(sandwich)
  library(lmtest)
  library(broom)
  library(patchwork)
  library(kableExtra)
}))

palette_grp <- c(
  "Common.Broadcaster" = "#0072B2", "Rare.Broadcaster" = "#56B4E9",
  "Common.Brooder"     = "#D55E00", "Rare.Brooder"     = "#E69F00"
)
label_grp <- c(
  "Common.Broadcaster" = "Common broadcaster", "Rare.Broadcaster" = "Rare broadcaster",
  "Common.Brooder"     = "Common brooder",     "Rare.Brooder"     = "Rare brooder"
)
```

## Step 1 — Read the octocoral survey data

The analysis begins from the colony survey file and the species-code table, both bundled collaborator data from the Lasker octocoral program (a standalone survey, not the RRS benthic cover section). The survey file holds one row per recorded colony: a census year, a site, a transect, a quadrat, a species code, and a living-tissue size class. The species-code table maps each code to a name and a reproductive mode.

```{r step1-read-octo}
#| code-summary: "Step 1 code: read the survey file and the species codes"
surveys <- read.csv("../../data/collab/octocoral_adultSurveys_Lasker_20250409.csv")
speciescodes <- read.csv("../../data/collab/octocoral_speciesCodes_Lasker_20250409.csv")
c(survey_rows = nrow(surveys), codes = nrow(speciescodes))
```

## Step 2 — Set the scope (publication range, sites, colonies)

The analysis keeps the columns it needs, marks each row as a counted colony, and holds the sampling design constant. The publication-range clip is the explicit first step: every downstream step runs only on years inside `pub_range`. At the published range this clip is harmless, since the bundled file does not extend beyond it; as the survey file grows in later monitoring years, the same clip is what keeps this page reproducing the published analysis exactly. After the clip, the analysis drops the sites and the one off-schedule survey that fall outside the comparison window, and marks a row as a colony when a living-tissue size class was recorded.

```{r step2-scope-octo}
#| code-summary: "Step 2 code: select columns, mark colonies, clip to pub_range, restrict sites"
octo <- surveys |>
  select(Census.Year, Site, Transect, Quadrat, Revised.Species.code, Living.tissue.size.class..1.9.9.0) |>
  rename(year = Census.Year, site = Site, transect = Transect, quadrat = Quadrat,
         coralSpecies = Revised.Species.code, sizeclass = Living.tissue.size.class..1.9.9.0) |>
  mutate(year = as.numeric(year),
         numcolonies = ifelse(sizeclass != "nd", 1, 0),
         coralSpecies = ifelse(coralSpecies == "#N/A", "nd", coralSpecies)) |>
  # Publication-range clip: the explicit first step of scope.
  filter(year >= pub_range[1], year <= pub_range[2]) |>
  filter(!site %in% c("Deep Tektite", "Yawzi", "Tektite"), year != 2017.11)
c(rows = nrow(octo), sites = n_distinct(octo$site), species_codes = n_distinct(octo$coralSpecies))
```

## Step 3 — Bin years and compute colony density

The octocoral record uses two-year groups for the baseline. Density is the analysis-ready measure: colonies counted, divided by the area of reef searched. This step counts colonies per species at each site and year, measures the area sampled from the number of quadrats searched, and divides to get colonies per square meter.

```{r step3-density-octo}
#| code-summary: "Step 3 code: bin years (two-year groups); compute colony density"
bin_years_octo <- function(df, interval = 2) {
  min_yr <- min(df$year, na.rm = TRUE); max_yr <- max(df$year, na.rm = TRUE)
  starts <- seq(min_yr, max_yr, by = interval)
  if (tail(starts, 1) + (interval - 1) > max_yr) starts <- head(starts, -1)
  map <- data.frame(year = seq(min_yr, max_yr)) |>
    rowwise() |>
    mutate(gs = max(starts[starts <= year]),
           year_group = paste(gs, min(gs + interval - 1, max_yr), sep = "-")) |>
    ungroup() |> select(year, year_group)
  df |> left_join(map, by = "year") |> mutate(year_group = as.factor(year_group))
}
octo <- bin_years_octo(octo, 2)
earliest_year_group <- levels(octo$year_group)[1]

# Colonies per species at each site and year.
colonies <- octo |>
  group_by(year_group, year, site, transect, quadrat, coralSpecies) |>
  summarise(numcolonies = sum(numcolonies, na.rm = TRUE), .groups = "drop")

# Area searched: the number of distinct quadrats per transect, summed per site and year.
area <- colonies |>
  group_by(year, site, transect) |> summarise(m2 = n_distinct(quadrat), .groups = "drop") |>
  group_by(year, site) |> summarise(m2_persite = sum(m2), .groups = "drop")

# Colony density (colonies per square meter), dropping unidentified colonies.
density <- colonies |>
  left_join(area, by = c("year", "site")) |>
  filter(coralSpecies != "nd") |>
  group_by(year_group, year, site, coralSpecies) |>
  summarise(ncols = sum(numcolonies, na.rm = TRUE), m2 = mean(m2_persite),
            coldens = sum(numcolonies, na.rm = TRUE) / mean(m2_persite), .groups = "drop")
c(rows = nrow(density), species = n_distinct(density$coralSpecies))
```

**Validation note:** the scoped dataset spans `r min(density$year)` through `r max(density$year)`, holds `r scales::comma(nrow(density))` species-site-year density records across `r n_distinct(density$site)` sites and `r n_distinct(density$coralSpecies)` octocoral species.

## Step 4 — Rank baseline abundance

Each species receives its mean colony density during the baseline period. Ranking the species by that mean sets the order used in the figure and sets up the commonness split. The cumulative proportion of density measures how much of the baseline density the top species account for.

```{r step4-baseline-octo}
#| code-summary: "Step 4 code: mean baseline density, rank, cumulative density"
baseline <- density |>
  filter(year_group == earliest_year_group) |>
  group_by(coralSpecies) |>
  summarise(meandens = mean(coldens), .groups = "drop") |>
  arrange(desc(meandens)) |>
  mutate(rank = row_number(),
         normdens = meandens / sum(meandens),
         cumsumdens = cumsum(normdens))
# head(baseline, 6)
```

## Step 5 — Classify common and rare (90/10)

The commonness split follows the same rule as the coral programs: the species that together hold the top 90 percent of baseline cumulative density are common, and the rest are rare.

```{r step5-classify-octo}
#| code-summary: "Step 5 code: split common from rare at 0.90 cumulative density"
baseline <- baseline |>
  mutate(commonness = ifelse(cumsumdens <= 0.90, "common", "Rare"))
table(baseline$commonness)
```

## Step 6 — Assign reproductive mode

Each species carries a reproductive mode, brooder or broadcaster, read from the species-code table. The table records the mode as a numeric code with a confidence level. This step maps the numeric codes to brooder or broadcaster, prefers a confirmed assignment over a likely one, and joins the mode onto the baseline species. Species without a mode leave the analysis.

```{r step6-repro-octo}
#| code-summary: "Step 6 code: map numeric codes to reproductive mode; join"
codes <- speciescodes |>
  filter(row_number() <= 69) |>
  transmute(coralSpecies = revised.species.code, coralSpeciesName = X,
            mode = case_when(reproductive.mode %in% c(1, 2) ~ "broadcaster",
                             reproductive.mode %in% c(3, 4) ~ "brooder", TRUE ~ NA_character_),
            confidence = case_when(reproductive.mode %in% c(1, 4) ~ "confirmed",
                                   reproductive.mode %in% c(2, 3) ~ "likely", TRUE ~ NA_character_)) |>
  filter(!is.na(mode)) |>
  arrange(coralSpecies, match(confidence, c("confirmed", "likely"))) |>
  group_by(coralSpecies) |> summarise(Reproductive_mode = first(mode),
                                       coralSpeciesName = first(coralSpeciesName), .groups = "drop")

baseline <- baseline |>
  left_join(codes, by = "coralSpecies") |>
  filter(Reproductive_mode %in% c("broadcaster", "brooder"))
# with(baseline, table(Commonness = commonness, `Reproductive mode` = Reproductive_mode))
```

## Step 7 — Build the analysis dataset

The model works on group means. This step joins the commonness and reproductive-mode labels onto the full density series, averages density within each group at each site and year, adds a year index that starts at zero, and takes the log10 of density.

```{r step7-dataset-octo}
#| code-summary: "Step 7 code: aggregate to group-site-year means"
analysis_data <- density |>
  inner_join(baseline |> select(coralSpecies, commonness, Reproductive_mode), by = "coralSpecies") |>
  group_by(category = commonness, reproductive_mode = Reproductive_mode, year, site) |>
  summarise(coldens = mean(coldens, na.rm = TRUE), .groups = "drop") |>
  mutate(yearind = year - min(year), coldensLog = log10(coldens)) |>
  filter(coldens > 0 & is.finite(coldensLog)) |>
  # Set explicit factor references (common, broadcaster) so the coefficient names match the
  # group-slope terms below. Without this the levels order by locale collation, which put
  # "Rare" before "common" here and broke the slope computation.
  mutate(category = factor(category, levels = c("common", "Rare")),
         reproductive_mode = factor(reproductive_mode, levels = c("broadcaster", "brooder")))
c(model_rows = nrow(analysis_data))
```

## Step 8 — Fit the model

The model predicts log10 colony density from year, commonness, reproductive mode, and every interaction among them. Common and broadcaster are the reference levels.

```{r step8-fit-octo}
#| code-summary: "Step 8 code: fit the three-way interaction model"
model_full <- lm(coldensLog ~ yearind * category * reproductive_mode, data = analysis_data)
```

## Step 9 — Robust standard errors

The residual variance differs across groups, so the analysis uses HC3 heteroscedasticity-consistent standard errors from a sandwich covariance matrix.

```{r step9-robust-octo}
#| code-summary: "Step 9 code: HC3 robust covariance and coefficient table"
vcov_hc3 <- sandwich::vcovHC(model_full, type = "HC3")
coef_robust <- broom::tidy(lmtest::coeftest(model_full, vcov = vcov_hc3))
```

## Step 10 — Group-specific slopes

The four group trends are linear combinations of the model coefficients, with standard errors from the robust covariance. Annual percent change back-transforms the log10 slope into a yearly rate.

```{r step10-slopes-octo}
#| code-summary: "Step 10 code: derive the four group slopes"
group_slope <- function(terms) {
  L <- setNames(rep(0, length(coef(model_full))), names(coef(model_full)))
  L[terms] <- 1
  est <- sum(L * coef(model_full))
  se <- sqrt(as.numeric(t(L) %*% vcov_hc3 %*% L))
  tibble(Slope = est, SE = se, `Annual % change` = 100 * (10^est - 1),
         `P-value` = 2 * pnorm(-abs(est / se)))
}
group_slopes <- bind_rows(
  "Common broadcaster" = group_slope("yearind"),
  "Rare broadcaster"   = group_slope(c("yearind", "yearind:categoryRare")),
  "Common brooder"     = group_slope(c("yearind", "yearind:reproductive_modebrooder")),
  "Rare brooder"       = group_slope(c("yearind", "yearind:categoryRare",
                                       "yearind:reproductive_modebrooder",
                                       "yearind:categoryRare:reproductive_modebrooder")),
  .id = "Group"
)
```

## Step 11 — Figure: abundance distribution and temporal trends

The figure pairs the baseline abundance distribution with the temporal trends. The left panel ranks species by baseline density and colors each by group. The right panel shows site-level group mean density over time on a log scale, with a linear fit per group.

```{r fig-octo}
#| code-summary: "Step 11 code: build the two-panel figure"
#| fig-width: 11
#| fig-height: 5.5
#| fig-cap: "Octocorals. Left: baseline abundance distribution, species ranked by mean baseline colony density, colored by commonness and reproductive mode. Right: site-level group mean colony density over time on a log scale, with per-group linear fits."
title_case <- function(x) recode(x, "common" = "Common", "Rare" = "Rare",
                                  "broadcaster" = "Broadcaster", "brooder" = "Brooder")
sad_df <- baseline |>
  filter(meandens > 0) |>
  mutate(coralSpecies = factor(coralSpecies, levels = rev(baseline$coralSpecies[order(baseline$rank)])),
         grp = interaction(title_case(commonness), title_case(Reproductive_mode)))
p_sad <- ggplot(sad_df, aes(coralSpecies, meandens, fill = grp)) +
  geom_col(width = 0.8) +
  scale_fill_manual(values = palette_grp, labels = label_grp, name = NULL, drop = FALSE) +
  coord_flip() +
  labs(x = NULL, y = "Mean baseline colony density", title = "Baseline abundance") +
  theme_classic(base_size = 11) +
  theme(legend.position = "top", axis.text.y = element_text(size = 7))

trend_df <- analysis_data |> mutate(grp = interaction(title_case(category), title_case(reproductive_mode)))
p_trend <- ggplot(trend_df, aes(year, coldens, color = grp, fill = grp)) +
  geom_jitter(shape = 21, color = "black", alpha = 0.2, width = 0.25, height = 0, show.legend = FALSE) +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE, alpha = 0.15, linewidth = 0.8) +
  scale_color_manual(values = palette_grp, labels = label_grp, name = NULL, drop = FALSE) +
  scale_fill_manual(values = palette_grp, labels = label_grp, name = NULL, drop = FALSE) +
  scale_y_log10() + annotation_logticks(sides = "l", color = "grey70") +
  labs(x = "Year", y = "Colony density (log scale)", title = "Temporal trends") +
  theme_classic(base_size = 11) +
  theme(legend.position = "top", axis.text.x = element_text(angle = 45, hjust = 1))

p_sad + p_trend
```

## Step 12 — Table: model coefficients

The table reports the model coefficients with HC3 robust standard errors and p-values. Bold marks terms significant at p less than 0.05.

```{r tbl-octo-coef}
#| code-summary: "Step 12 code: format the coefficient table and the group slopes"
term_labels <- c(
  "(Intercept)" = "Intercept", "yearind" = "Year",
  "categoryRare" = "Rarity", "reproductive_modebrooder" = "Repro. mode (brooder)",
  "yearind:categoryRare" = "Year x Rarity", "yearind:reproductive_modebrooder" = "Year x Repro. mode",
  "categoryRare:reproductive_modebrooder" = "Rarity x Repro. mode",
  "yearind:categoryRare:reproductive_modebrooder" = "Year x Rarity x Repro. mode"
)
coef_tbl <- coef_robust |>
  transmute(Predictor = term_labels[term],
            `Estimate (SE)` = sprintf("%.3f (%.3f)", estimate, std.error),
            `P-value` = ifelse(p.value < 0.001, "< 0.001", sprintf("%.3f", p.value)),
            .sig = p.value < 0.05)

kbl(coef_tbl |> select(-.sig), align = c("l", "r", "r"),
    caption = "Octocoral model coefficients with HC3 robust standard errors (log10 colony density).") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) |>
  row_spec(which(coef_tbl$.sig), bold = TRUE)
```

```{r tbl-octo-slopes}
#| code-summary: "Step 12 code: the four group slopes"
group_slopes |>
  mutate(Slope = sprintf("%.4f", Slope), SE = sprintf("%.4f", SE),
         `Annual % change` = sprintf("%.2f", `Annual % change`),
         `P-value` = ifelse(`P-value` < 0.001, "< 0.001", sprintf("%.3f", `P-value`))) |>
  kbl(align = c("l", "r", "r", "r", "r"),
      caption = "Group-specific temporal slopes (octocorals), with robust SEs and annual percent change.") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

## Step 13 — Save the analysis objects

The octocoral analysis keeps the same abundance table it already builds above. Like the coral programs, it now also writes its analysis objects to a timestamped RData file under `data/rdata/`, tagged with the program and the version (as published or updated) so the two ranges never overwrite each other. The manuscript-values page reads the octocoral study window and the octocoral group slopes from this saved file.

```{r save-rdata-octo}
#| code-summary: "Step 13 code: save the key objects to a timestamped RData"
#| include: false
source("../../_includes/_analysis_helpers.R")
if (!exists("version")) version <- "unversioned"
rdata_file <- save_program_rdata(program, version, list(
  baseline = baseline, analysis_data = analysis_data, model_full = model_full,
  vcov_hc3 = vcov_hc3, coef_robust = coef_robust, group_slopes = group_slopes,
  species_table = baseline, pub_range = pub_range, program = program,
  taxon_level = taxon_level, version = version, earliest_year_group = earliest_year_group
))
```

## Step 14 — SI figure: model diagnostics

The octocoral coefficient table uses the same HC3 robust standard errors as the coral programs. These diagnostic panels show the residual structure that motivates them: residuals versus fitted values, a normal Q-Q plot of standardized residuals, and the residual distribution.

```{r sfig-octo-diagnostics}
#| code-summary: "Step 14 code: model diagnostics from the fitted model"
#| fig-width: 11
#| fig-height: 3.5
#| fig-cap: "Octocoral model diagnostics: residuals versus fitted values, a normal Q-Q plot of standardized residuals, and the residual distribution."
model_diagnostics_plot(model_full)
```
